The natural history of scavenging in vertebrates

### 2016-03-22 natural history of scavenging.tex
Added vultures age
added Graeme's ideas

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers \citep{tenney1877few}. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature. Fortunately, research on scavenging is on the rise \citep{manga2006vulture}. As a result, we are now beginning to realise the extent of this behaviour such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{benbow2015introduction}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses. By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper. Here we chart the natural history of vertebrate scavenging on air, land and in water by focussing on the dominant vertebrate groups. %It is our aim in this review to employ these methods to gain an understanding of scavengers past and present. 
\subsection*{The Difficulty of Scavenging}
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Once found the scavenger has to then overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}. The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}. Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}. The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

\section*{Aerial Scavengers}

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Flight is also a cheaper means of locomotion than running \citep{tucker1975energetic}. This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}. The benefits this confers are clear from the information we have on the enormous foraging ranges of some seabirds and accipiters \citep{bamford2007ranging}. Clearly, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it. Although, as with detection ability, the environment has a role to play here. Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}. In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist. The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.

Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya). Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres \citep{witton2010size})%CITE
 and look to have engaged in soaring flight \citep{witton2010size}. Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

@@

\section*{Terrestrial Scavengers}
A simplification of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. The same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

Terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet. In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be as high as 99\% \citep{benbow2015introduction}. Therefore, we can again use these species as our efficient terrestrial scavengers to compare with other forms.

Similar to vultures they have well developed sensory organs, particularly in olfaction whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait"  which allows them to cover great distances efficiently. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores. Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.

@@

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. 
%TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.
The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment.

By contrast to mammals, a successful reptilian scavenger requires a far different set of adapations. Modern forms are ectothermic, limiting their activity periods. This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}. A lower metabolism does give reptiles an advantage however, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}. Any adaptations that reduce energetic costs are likely to be selected in scavengers. \cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they "exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year. It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is "far more common than currently acknowledged."\citep{devault2002scavenging}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}. For the smaller species, larger competitors would have prevented their access to carrion.

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex}. There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet. As noted earlier, a system similar to the Serengeti in productivity could have supported a terrestrial scavenger \citep{ruxton2004obligate}.

\section*{Aquatic Scavengers}

An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In the latter low-light environment, detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

The presence of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if a scavenger could survive by seeking out these remains exclusively. \cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely. Any animal that could seek out such whale carcasses is unlikely to have ignored other types of carrion. Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of whales, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1-163.5 Mya), that weighed in excess of 20 tonnes. Thus, the energetic feasiblity of a marine scavenger has a long history. One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}. This meant Killer Whales \textit{Orcinus orca} could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the "dinner-bells" \citep{Whitehead415}. Early whales such as \textit{Basilosaurus} seem to fit into the same niche as Killer Whales and we have some evidence for scavenging in this group as well \citep{fahlke2012bite}.

\section*{Ecological Role}
It is recognised that scavengers keep energy flows at a higher trophic level in food webs than decomposers because they consume relatively more carrion \citep{devault2003scavenging}. They are also hugely important for the dispersal of nutrients \citep{benbow2015introduction}. Consider the diversity of animals that can end up feeding at the carcass of an elephant. Here we have an incredibly dense and nutrient rich patch that ends up being distributed widely. In the absence of vertebrate scavengers, invertebrates and microorganisms would consume the carcass in-situ or at least distribute the constituent nutrients over a much shorter range. This effect has been magnified as vertebrates evolved certain key traits that allowed them to range farther, namely an upright gait, an endothermic metabolism and of course, flight. To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass. In the case of mammals and reptiles these are 1.15 * body mass (kg) \textsuperscript{0.12} and 0.23 * body mass (kg) \textsuperscript{0.12}
respectively \citep{ruxton2004obligate}. We can insert these into a foraging radius model ((duration * speed)/2)/1000 for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. Thus, in an ecological context, the evolution of these steps coupled with the ability to scavenge resulted in a world with a far more widely distributed nutrient landscape.

\section*{The Role of Climate}
The geological record shows the Earth has undergone radical fluctuations in temperature. This will have had a significant bearing on the availability of carrion. A 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{benbow2015introduction}. Oceanic productivity, decomposition rates and habitat structure are all impacted by climactic conditions. \cite{benbow2015introduction} points to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers "above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses". Furthermore, "carcass decomposition occurs more rapidly in warm, moist habitats when compared with more xeric habitats."

>>>>

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers \citep{tenney1877few}. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature. Fortunately, research on scavenging is on the rise \citep{manga2006vulture}. As a result, we are now beginning to realise the extent of this behaviour such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{beasley2015vertebrates}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses. By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper. Here we chart the natural history of vertebrate scavenging on air, land and in water by focussing on the dominant vertebrate groups that existed at each time. %It is our aim in this review to employ these methods to gain an understanding of scavengers past and present. 
\subsection*{The Difficulty of Scavenging}
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Once found, the scavenger has to then overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}. The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}. Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}. The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

\section*{Aerial Scavengers}
Scavenging should be particularly attractive to avian predators compared to mammals. Solitary mammalian predators can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier (e.g. socially hunting lions \citep{owen2008predator}) . In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey}. This is likely due to their need to kill prey that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal. Thus for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene (66 - 56 Mya) \citep{Jetz2012, Jarvis2014}. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Flight is also a cheaper means of locomotion than running \citep{tucker1975energetic}. This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}. The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors}. Clearly, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it. Although, as with detection ability, the environment has a role to play here. Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}. In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist. The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.

Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya). Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres \citep{witton2010size}) and look to have engaged in soaring flight \citep{witton2010size}. Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

@@

\section*{Terrestrial Scavengers}
A simplification of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent \citep{beasley2015vertebrates}. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. Interestingly, the same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

Terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet. In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}. Therefore, we can again use these species as our efficient terrestrial scavengers to compare with other forms.

Similar to vultures they have well developed sensory organs, whereby they can smell a rotting carcass 4 km away and can hear the vocalisations of conspecifics at a distance of 10 km \citep{mills1989comparative}. They have a characteristic "rocking horse gait"  which allows them to cover great distances efficiently \citep{jones2015african}. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores and given some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result, the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.

@@

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas are likely to have left far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. 
%TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.
The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment \citep{moleon2014humans}.

By contrast to mammals, a successful reptilian scavenger requires a far different set of adapations. Modern forms are ectothermic, limiting their activity periods. This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}. A lower metabolism does give reptiles an advantage however, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}. Any adaptations that reduce energetic costs are likely to be selected in scavengers. \cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they "exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year. It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is "far more common than currently acknowledged."\citep{devault2002scavenging}. The same reasoning can be applied to crocodlies and their allies \citep{forrest2003evidence}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception \citep{farlow1994speculations}. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}. For the smaller species, larger competitors would have prevented their access to carrion.

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex} \citep{hone2010feeding}. There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet. As noted earlier, a system similar to the Serengeti in productivity could have supported a terrestrial scavenger \citep{ruxton2004obligate}. Indeed, in systems that were dominated by large ectothermic or mesothermic vertebrates 
the same primary productivity would have supported a greater biomass \citep{mcnab2009resources}. The upshot of this is there was a higher biomass of herbivores dying and offering scavenging opportunities. Predators were large bodied too compared to extant mammalian predators \citep{mcnab2009resources}, and so, especially if they were ectothermic, could last longer between meals rendering scavenging a more attractive behaviour relative to predation.

\section*{Aquatic Scavengers} 
Primary productivity is lower in almost all aquatic systems than terrestrial systems (except deserts) so as we go up the food chain the density of carcasses worth scavenging is going to be lower. But it is certainly known to occur in many aquatic vertebrates. An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although, as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark, \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.  
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species, a carcass in this environment either floats or descends to the sea floor \citep{Whitehead415}. In the latter low-light environment, visual detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{beasley2015vertebrates} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

The presence of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if a scavenger could survive by seeking out these remains exclusively. \cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely. Any animal that could seek out such whale carcasses is unlikely to have ignored other types of carrion. Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of whales, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1-163.5 Mya), that weighed in excess of 20 tonnes. Thus, the energetic feasiblity of a marine scavenger has a long history. One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}. This meant Killer Whales (\textit{Orcinus orca}) could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the "dinner-bells" \citep{Whitehead415}. Early whales such as \textit{Basilosaurus} seem to fit into the same niche as Killer Whales and we have some evidence for scavenging in this group as well \citep{fahlke2012bite}. A final point to make is that vertebrates are relatively rarer in aquatic environments, because even large animals can get support from the buoyancy of the water without needing a backbone.

\section*{Ecological Role}
It is recognised that scavengers keep energy flows at a higher trophic level in food webs than decomposers because they consume relatively more carrion \citep{devault2003scavenging}. They are also hugely important for the dispersal of nutrients \citep{beasley2015vertebrates}. Consider the diversity of animals that can end up feeding at the carcass of an elephant. Here we have an incredibly dense and nutrient rich patch that ends up being distributed widely. In the absence of vertebrate scavengers, invertebrates and microorganisms would consume the carcass in-situ or at least distribute the constituent nutrients over a much shorter range. This effect has been magnified as vertebrates evolved certain key traits that allowed them to range farther, namely an upright gait, an endothermic metabolism and of course, flight. To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass. In the case of mammals and reptiles these are 1.15 * body mass (kg) \textsuperscript{0.12} and 0.23 * body mass (kg) \textsuperscript{0.12}
respectively \citep{ruxton2004obligate}. We can insert these into a foraging radius model ((duration * speed)/2)/1000 for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. Thus, in an ecological context, the evolution of these steps coupled with the ability to scavenge resulted in a world with a far more widely distributed nutrient landscape.
